A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
more

@@

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables.
For example, 
%$\mu x. \const{C}( x ) \aequiv \mu y. \const{C}( y )$ and
$\mu x.\; \const{D}( y, x ) \aequiv \mu z.\; \const{D}( y, z )$,
but
$\mu x.\; \const{C}( x ) \not\aequiv \mu x.\; \const{D}( y, x ) \not\aequiv \mu x.\; \const{D}( z, x )
\not\aequiv \mu y.\; \const{D}( y, x )$.
Two $\mu$-terms can represent the same value while not being $\alpha$-equivalent---e.g.,
$\mu x.\; \const{C}( x ) \not\aequiv \mu x.\; \const{C}( \const{C}( x ) )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable $\Var{\ec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = \Var{\ec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following rule exhaustively:
\[
\hbox{\(
\inferrule{
  \Var{\ec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ \Var{\ec{x}} \mapsto \mu \Var{\ec{x}}.\; \const{C}( \Var{\ec{t_1}}, \ldots, \Var{\ec{t_n}} ) \}
}
\)
\rn{Refine}}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

\begin{example}
Say that $\Ec$ contains (distinct) equivalence classes $\ec{x}$, $\ec{y}$, $\ec{z}$ and $\ec{w}$ 
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of updates applied to $\Val$ is shown below.
In the first column, we show the equivalence class that was considered,
and in the second column, we show the content of $\Val$.
\[\begin{array}{|l|@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l|}
\hline
\text{Eqc} & \multicolumn{3}{c}{\Val} & \ 
\\
\hline
& \{ & \ec{x} \mapsto \Var{\ec{x}}, & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \Var{\ec{y}}, & \ec{w} \mapsto \Var{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \Var{\ec{y}} ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \Var{\ec{y}}, & \ec{w} \mapsto \Var{\ec{w}} & \} \\
\ec{y}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ) ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ), & \ec{w} \mapsto \Var{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ) ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \Var{\ec{y}} ) ), & \ec{w} \mapsto \Var{\ec{w}} & \} \\
\hline
\end{array}\]
This indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( \Var{\ec{w}}, \const{C}( \Var{\ec{z}}, \ldots ))$ (resp. $\const{C}( \Var{\ec{z}}, \const{C}( \Var{\ec{w}}, \ldots ))$) 
for some value of $\Var{\ec{w}}$ and $\Var{\ec{z}}$.
\qed
\end{example}

@@

\begin{table*}[tbh!]
\normalsize
\begin{center}\begin{tabular}{l@{\kern1.5em}c@{\kern.5em}c@{\kern.5em}c@{\kern1.5em}c}
  & \HD{Distro} & \HD{AFP} & \HD{\gandl} & All
\\[1pt] %%% TYPESETTING HACK
\midrule
%$\enatT$
\\[-9pt] %%% TYPESETTING HACK
No (co)datatypes
  & 221 & 760 & \phantom{0}57 & 1038 \\
Datatypes without acyclicity
  & 227 & 765 & \phantom{0}57 & 1049 \\
Datatypes with acyclicity
  & 227 & 771 & \phantom{0}57 & 1055 \\
Codatatypes without bisimilarity
  & 217 & 787 & \phantom{0}55 & 1059 \\
Codatatypes with bisimilarity
  & 218 & 787 & \phantom{0}58 & 1063 \\
Full (co)datatypes
  & 224 & 798 & \phantom{0}58 & 1080 \\[\jot]
Total number of goals
  & 879 & 2974\phantom{0} & 317 & 4170
\end{tabular}\end{center}
\caption{\,Number of solved goals for the three benchmark suites}
\label{tab:bench}
\end{table*}

>>>>

\newcommand\MIDRULE{
\\[-1pt] %%% TYPESETTING HACK
\midrule
%$\enatT$
\\[-11pt] %%% TYPESETTING HACK
}

@@

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables---e.g.,
%$\mu x. \const{C}( x ) \aequiv \mu y. \const{C}( y )$ and
$\mu x.\; \const{D}( y, x ) \aequiv \mu z.\; \const{D}( y, z )$\begin{report},
but
$\mu x.\; \const{C}( x ) \not\aequiv \mu x.\; \const{D}( y, x ) \not\aequiv \mu x.\; \const{D}( z, x )
\not\aequiv \mu y.\; \const{D}( y, x )$\end{report}.
Two $\mu$-terms can represent the same value while not being $\alpha$-equivalent---e.g.,
$\mu x.\; \const{C}( x ) \not\aequiv \mu x.\; \const{C}( \const{C}( x ) )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

The mapping $\Val$ is computed as follows.
%
For each equivalence class $\ec{x}$, we associate a fresh variable $\Var{\ec{x}}$ of the same type as $x$
%not occurring in $\tEc$,
and let $\Val\ec{x} = \Var{\ec{x}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{x}$.
The mapping $\Val$ is then modified by applying the following rule exhaustively:
\[
\hbox{\(
\inferrule{
  \Var{\ec{x}} \in \FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ \Var{\ec{x}} \mapsto \mu \Var{\ec{x}}.\; \const{C}( \Var{\ec{t_1}}, \ldots, \Var{\ec{t_n}} ) \}
}
\)
\rn{Update}}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Update} rule applications is presented below.
\[\begin{array}{l@{\hspace{1em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
%\hline
%\text{Step} & \multicolumn{3}{l}{\text{Mapping }\Val}
%\MIDRULE
\text{1. Initial:}
& \{ & \ec{x} \mapsto \Var{\ec{x}}, & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \Var{\ec{y}}, & \ec{w} \mapsto \Var{\ec{w}} & \} \\[\jot]
\text{2. Update w.r.t.\ }\ec{x}\text{:}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \Var{\ec{y}} ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \Var{\ec{y}}, & \ec{w} \mapsto \Var{\ec{w}} & \} \\[\jot]
\text{3. Update w.r.t.\ }\ec{y}\text{:}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ) ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ), & \ec{w} \mapsto \Var{\ec{w}} & \} \\[\jot]
\text{4. Update w.r.t.\ }\ec{x}\text{:}
& \{ & \ec{x} \mapsto \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \Var{\ec{x}} ) ), & \ec{z} \mapsto \Var{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu \Var{\ec{y}}. \const{C}( \Var{\ec{z}}, \mu \Var{\ec{x}}. \const{C}( \Var{\ec{w}}, \Var{\ec{y}} ) ), & \ec{w} \mapsto \Var{\ec{w}} & \}
%\hline
\end{array}\]
%The first column shows the equivalence class that was considered;
%the second column shows the content of $\Val$.
This indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( \Var{\ec{w}}, \const{C}( \Var{\ec{z}}, \ldots ))$ (resp. $\const{C}( \Var{\ec{z}}, \const{C}( \Var{\ec{w}}, \ldots ))$) 
for some value of $\Var{\ec{w}}$ and $\Var{\ec{z}}$.
\qed
\end{example}

@@

\begin{table*}[tbh!]
\normalsize
\begin{center}\begin{tabular}{l@{\kern1.5em}c@{\kern.5em}c@{\kern.5em}c@{\kern1.5em}c}
  & \HD{Distro} & \HD{AFP} & \HD{\gandl} & All
\MIDRULE
No (co)datatypes
  & 221 & 760 & \phantom{0}57 & 1038 \\
Datatypes without acyclicity
  & 227 & 765 & \phantom{0}57 & 1049 \\
Datatypes with acyclicity
  & 227 & 771 & \phantom{0}57 & 1055 \\
Codatatypes without bisimilarity
  & 217 & 787 & \phantom{0}55 & 1059 \\
Codatatypes with bisimilarity
  & 218 & 787 & \phantom{0}58 & 1063 \\
Full (co)datatypes
  & 224 & 798 & \phantom{0}58 & 1080 \\[\jot]
Total number of goals
  & 879 & 2974\phantom{0} & 317 & 4170
\end{tabular}\end{center}
\caption{\,Number of solved goals for the three benchmark suites}
\label{tab:bench}
\end{table*}
